---
title: "Introduction to R and RStudio"
output: statsr:::statswithr_lab
---

<div id="instructions">
Complete all **Exercises**, and submit answers to **Questions** on the Coursera 
platform.
</div>

## RStudio

Your RStudio window has four panels.

Your R Markdown file (this document) is in the upper left panel.

The panel on the lower left is where the action happens. It's called the *console*. 
Everytime you launch RStudio, it will have the same text at the top of the 
console telling you the version of R that you're running. Below that information
is the *prompt*. As its name suggests, this prompt is really a request, a 
request for a command. Initially, interacting with R is all about typing commands
and interpreting the output. These commands and their syntax have evolved over
decades (literally) and now provide what many users feel is a fairly natural way
to access data and organize, describe, and invoke statistical computations.

The panel in the upper right contains your *workspace* as well as a history of 
the commands that you've previously entered. 

Any plots that you generate will show up in the panel in the lower right corner. 
This is also where you can browse your files, access help, manage packages, etc.

## R Packages

R is an open-source programming language, meaning that users can contribute
packages that make our lives easier, and we can use them for free. For this lab,
and many others in the future, we will use the following R packages:

- statsr: for data files and functions used in this course
- dplyr: for data wrangling
- ggplot2: for data visualization

You should have already installed these packages using commands like 
install.packages and install_github.

Next, you need to load the packages in your working environment. We do this with
the library function. Note that you only need to **install** packages once, but
you need to **load** them each time you relaunch RStudio.

```{r load-packages, message = FALSE}
library(dplyr)
library(ggplot2)
library(statsr)
```

## Dataset 1: Dr. Arbuthnot's Baptism Records

```{r load-abruthnot-data}
data(arbuthnot)
```

```{r view-data}
arbuthnot
```

```{r dim-data}
dim(arbuthnot)
```

```{r names-data}
names(arbuthnot)
```

1. How many variables are included in this data set?

The dataset includes **3** variables: year, boys, and girls.

**Exercise**: What years are included in this dataset? 

```{r}
range(arbuthnot$year)
```

The years included in the dataset range from 1629 to 1710.

2. What command would you use to extract just the counts of girls born? 

The command to extract the counts of girls born is `arbuthnot$girls`.

```{r extract-counts-of-girls-born}
arbuthnot$girls
```

The command correctly extracts the number of girls baptized each year.

**Exercise**: Generate a plot of the proportion of boys born over time.

```{r plot-proportion-of-boys-over-time}
arbuthnot <- arbuthnot %>%
  mutate(proportion_boys = boys / (boys + girls))

ggplot(data = arbuthnot, aes(x = year, y = proportion_boys)) +
  geom_line() +
  geom_point() +
  labs(y = "Proportion of Boys", x = "Year")
```

## Dataset 2: Present birth records

```{r load-present-data}
data(present)
```

4. How many variables are included in this data set?

```{r variables-in-present}
dim(present)
```

The dataset includes **4** variables.


**Exercise**: What years are included in this dataset?

```{r years-in-present-data}
range(present$year)
```

The years included in the dataset range from 1940 to 2013.

5. Calculate the total number of births for each year and store these values in a new variable called `total`. Then, calculate the proportion of boys born each year and store these values in a new variable called `prop_boys` in the same dataset. Plot these values over time and determine if the proportion of boys born in the US has decreased over time.

```{r prop-boys-over-time}
present <- present %>%
  mutate(total = boys + girls,
         prop_boys = boys / total)

ggplot(data = present, aes(x = year, y = prop_boys)) +
  geom_line() +
  geom_point() +
  labs(y = "Proportion of Boys", x = "Year")
```

The plot shows the proportion of boys born each year. Based on the plot, the statement that the proportion of boys born in the US has decreased over time is **False**.


6. Create a new variable called `more_boys` which contains TRUE if that year had more boys than girls, or FALSE if that year did not. Based on this variable, which of the following statements is true?

```{r more-boys-per-year}
present <- present %>%
  mutate(more_boys = boys > girls)

table(present$more_boys)
```


The variable `more_boys` indicates whether there were more boys born than girls in each year. Based on this variable, the correct statement is: **Half of the years there are more boys born, and the other half more girls born**.


7. Calculate the boy-to-girl ratio each year and store these values in a new variable called `prop_boy_girl`. Plot these values over time. Which of the following best describes the trend?

```{r prop-boy-girl-over-time}
present <- present %>%
  mutate(prop_boy_girl = boys / girls)

ggplot(data = present, aes(x = year, y = prop_boy_girl)) +
  geom_line() +
  geom_point() +
  labs(y = "Boy-to-Girl Ratio", x = "Year")
```


The plot of the boy-to-girl ratio over time suggests that **there is initially an increase in boy-to-girl ratio, which peaks around 1960. After 1960 there is a decrease in the boy-to-girl ratio, but the number begins to increase in the mid 1970s**.


8. In what year did we see the most total number of births in the U.S.?

```{r most-total-births}
present %>%
  mutate(total = boys + girls) %>%
  arrange(desc(total)) %>%
  slice(1)
```

The year with the most total number of births in the U.S. was **1991**.


## Resources for learning R and working in RStudio

That was a short introduction to R and RStudio, but we will provide you with more
functions and a more complete sense of the language as the course progresses. You 
might find the following tips and resources helpful.

- In this course we will be using the dplyr (for data wrangling) and ggplot2 (for 
data visualization) extensively. If you are googling for R code, make sure
to also include these package names in your search query. For example, instead
of googling "scatterplot in R", google "scatterplot in R with ggplot2".

- The following cheathseets may come in handy throughout the course. Note that some 
of the code on these cheatsheets may be too advanced for this course, however 
majority of it will become useful as you progress through the course material.
    - [Data wrangling cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
    - [Data visualization cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)
    - [R Markdown](http://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)

- While you will get plenty of exercise working with these packages in the labs of 
this course, if you would like further opportunities to practice we recommend 
checking out the relevant courses at [DataCamp](https://www.datacamp.com/courses).

<div id="license">
This is a derivative of an [OpenIntro](https://www.openintro.org/stat/labs.php) lab, and is released under a [Attribution-NonCommercial-ShareAlike 3.0 United States](https://creativecommons.org/licenses/by-nc-sa/3.0/us/) license.
</div>
